Higher-Order unification for free!

### 2024-04-17 paper/main.tex
cite beluga

\begin{abstract}
  Specifying and implementing a proof system from scratch requires significant effort.
  Logical Frameworks and Higher Order Logic Programming Languages provide
  dedicated, high-level meta languages to facilitate this task in two
  % key 
  ways: 1) variable binding and substitution are for free when meta language binders
  represent object logic ones; 2) proof construction, and %even 
  proof search, are
  greatly simplified by leveraging the unification procedure provided by the meta language.
  Notable examples of meta languages are Elf~\cite{elf}, Twelf~\cite{twelf},
  $\lambda$Prolog~\cite{miller_nadathur_2012} and
  Isabelle~\cite{10.1007/978-3-540-71067-7_7}
  which have been %utilized
  used to implement %various
  many formal systems such as
  First Order Logic~\cite{felty88cade},
  Set Theory~\cite{10.1007/BF00881873},
  Higher Order Logic~\cite{books/sp/NipkowPW02}, and %even
  the Calculus of
  Constructions~\cite{felty93lics}.

  The object logic we are interested in is
  Coq's Calculus of Inductive Constructions (CIC)\cite{Coq-refman}. We aim to
  develop a higher-order unification-based 
  proof search procedure for it using the meta language
  Elpi~\cite{dunchev15lpar}, a dialect of $\lambda$Prolog.
  Elpi's equational theory includes
  $\eta\beta$-equivalence and features a
  higher-order unification procedure \Ue restricted to the pattern
  fragment~\cite{miller92jsc}.
  Elpi offers an encoding of CIC suitable
  for meta-programming~\cite{tassi:hal-01637063,tassi:hal-01897468,gregoire:hal-03800154,newtc}
  but restricts \Ue to roughly first-order unification problems only.
  We refer to this basic encoding as \Fo.

  In this paper we propose a more well-behaved encoding called \Ho,
  and show how to translate unification problems from \Fo{}
  to corresponding ones in \Ho.
  Consequently, we derive \Uo{}, the higher-order unification procedure of \Fo{}
    that honours $\eta\beta$-equivalence (for CIC functions), addresses
  problems within the pattern fragment, and allows
  for the use of heuristics to deal with problems outside the
  pattern fragment. Moreover, as \Uo{} delegates most of the work to \Ue,
  it can be used to efficiently simulate a logic program in \Fo{} by
  taking advantage of
  unification-related optimizations of the meta language, such as clause indexing.

>>>>

\begin{abstract}
  Specifying and implementing a proof system from scratch requires significant effort.
  Logical Frameworks and Higher Order Logic Programming Languages provide
  dedicated, high-level meta languages to facilitate this task in two
  % key 
  ways: 1) variable binding and substitution are for free when meta language binders
  represent object logic ones; 2) proof construction, and %even 
  proof search, are
  greatly simplified by leveraging the unification procedure provided by the meta language.
  Notable examples of meta languages are Twelf~\cite{twelf},
  $\lambda$Prolog~\cite{miller_nadathur_2012}, Beluga~\cite{Beluga} and
  Isabelle~\cite{10.1007/978-3-540-71067-7_7}
  which have been %utilized
  used to implement or specify %various
  many
  formal systems such as
  First Order Logic~\cite{felty88cade},
  Set Theory~\cite{10.1007/BF00881873},
  Higher Order Logic~\cite{books/sp/NipkowPW02}, and %even
  the Calculus of
  Constructions~\cite{felty93lics}.

  The object logic we are interested in is
  Coq's type theory~\cite{Coq-refman}. We aim to
  develop a higher-order unification-based 
  proof search procedure %for it 
  using the meta language
  Elpi~\cite{dunchev15lpar}, a dialect of $\lambda$Prolog.
  Elpi's equational theory includes
  $\eta\beta$-equivalence and features a
  higher-order unification procedure \Ue 
  %restricted to the
  for the
  pattern fragment~\cite{miller92jsc}.
  Elpi offers an encoding of Coq terms suitable
  for meta programming~\cite{tassi:hal-01637063,tassi:hal-01897468,gregoire:hal-03800154,newtc}
  but that restricts \Ue to first-order unification problems only.
  We refer to this basic encoding as \Fo.

  In this paper we propose a more well-behaved encoding called \Ho,
  and show how to translate unification problems from \Fo{}
  to corresponding ones in \Ho.
  Consequently, we derive \Uo{}, the higher-order unification procedure of \Fo{}
    that honours $\eta\beta$-equivalence (for Coq functions), addresses
  problems within the pattern fragment, and allows
  for the use of heuristics to deal with problems outside the
  pattern fragment. Moreover, as \Uo{} delegates most of the work to \Ue,
  it can be used to efficiently simulate a logic program in \Fo{} by
  taking advantage of
  unification-related optimizations of the meta language, such as clause indexing.
